Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
synth qre lenses

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption,
cleveref}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\usepackage{xspace}

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\codefont{#1}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

Primitive lenses include:
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: insert the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
Such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, as well as lens composition.
Figure~\ref{fig:example-lens} illustrates the use of these
combinators to define a lens \cd{bib_to_end} that transforms data
in \bibtex{} to EndNote (and vice versa). 
\begin{figure}[t]
\begin{lstlisting}
let preamble : (lens in "@book{" $\Leftrightarrow$ "%0 Book\n%F ") =
del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
del ",\nauthor = \""
. ins "\n%A "
. name
. (del " and "
. ins "\n"
. (ins "%A " . name . del " and " . ins "\n")*
. ins "%A "
. name
|| "")

let title_lens : (lens in ("\",\n" . bib_title . "},\n}") $\Leftrightarrow$ ("\n" . end_title)) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : (lens in bibtex $\Leftrightarrow$ end_note) = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\ifdraft
\bcp{And a readability suggestion:
  define a named constant for a quote character.}
\fi}
\label{fig:example-lens}
\end{figure}

@@

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\ksf{What is the notation ``lens in'' below?  We should explain it
intuitively here.}
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.\saz{Why does the ``cref'' macro not capitalize Section?}

\subsection{Syntax of Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R}{R'}{f} \bsep{} \id(R) \bsep \project{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, each QRE $Q$ enables us to express four semantic objects

\begin{tabular}{rl}
  $W(Q)$  & a regular expression (the ``whole'' of $Q$),\\
  $\eqrel{Q}$  & an equivalence relation on $\mathcal{L}(W(Q))$,\\
  $K(Q)$ & a regular expression (the ``kernel'' of $Q$) such that
           $\mathcal{L}(K(Q))$ forms a complete \\
             & set of representatives for $\eqrel{Q}$, and \\
  $\canonize(Q)$ & 
  a ``canonizing'' function that, given any $w \in \mathcal{L}(W(Q))$,
  yields $\canonize(Q)(w)$ \\ & as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

The well-formedness constraints for QREs ensure that these four semantic
components fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by \canonize(Q).

\subsection{The \kw{normalize} Combinator}

The relationship among the semantic objects of a QRE can be understood in terms
of the \normalize{R}{S}{f} combinator, which exposes each of these pieces
explicitly.  Its whole language is just $R$, its kernel language is just
$S$, and $\canonize(\normalize{R}{S}{f}) = f$.  Its equivalence relation is
determined by the fibres of $f$, so we have
$s_1 \equiv s_2 \Leftrightarrow f(s_1) = f(s_2)$.

These components form a coherent quotient language when the canonization
function $f$ is surjective and idempotent (intuitively, $f$ picks out a unique
representative for each equivalence class).  We also require that the kernel
language be a subset of the whole language, which enables QRE composition.
These considerations lead to the following well-formedness rule.

\begin{prooftree}
\AxiomC{$\mathcal{L}(S) \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.   However,
although \kw{normalize} is powerful, verifying that a canonization function $f$
is surjective and idempotent is in general undecidable.  This means that a
programmer wishing to use \kw{normalize} must discharge strong proof
obligations, which is cumbersome in practice.\footnote{In our implementation we allow a programmer to use \kw{normalize} at their
own risk without checking these side conditions as an ``escape hatch'' for the
case when other QRE combinators are insufficient.}

The remaining QRE combinators provide simpler, more compositional ways of
building canonizers.  Nevertheless, the \kw{normalize} combinator gives a
sufficient condition for potential QREs to be well-behaved, particularly with
respect to the strategy for synthesizing quotient lenses which we shall
introduce in Section~\ref{synth}, so we use it to guide the design of our QRE
combinators, as we see next.

% ; this is the second
% property that distinguishes the \kw{normalize} combinator from the other QREs.
% The \kw{normalize} combinator is special in two main ways, the first of 
% which has to do with the premises in its inference rule:

\subsection{QRE Combinator Semantics}

Figure~\ref{fig:wk} gives the inductive definitions of the whole and
kernel languages $W(Q)$ and $K(Q)$ for all of the QRE forms.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \kw{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \kw{permutation} combinator allows for the string to match any
permutation the $Q_i$'s while also taking into account the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \kw{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \kw{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

@@

The most complicated inference rule is the (Perm) rule for the
\kw{permutation} combinator. The second hypothesis for the
Perm rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\kw{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \kw{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$ which is
the kernel of the \kw{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

\begin{lstlisting}
let bib_to_end_q : (lens in bibtex $\Leftrightarrow$ endnote) =  rquot (lquot(bibtex, bib_to_end), endnote)
\end{lstlisting}

@@

The Optician synthesis algorithm works as follows. Given regular expressions $R$
and $S$ and a set of example input--output pairs
$\{(r_1, s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to
find a bijective lens $\ell : R \Leftrightarrow S$ that that agrees on the
examples, \textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and
$s_i$ to $r_i$ in the backward direction.  The Optician algorithm is guaranteed
to succeed (eventually!) if such a lens exists.    \saz{We should probably
  mention what happens if the examples aren't sufficient to uniquely
  determine the lens.}

>>>>

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{algorithm, amsmath, amssymb, verbatim, enumerate, graphicx,
  centernot, tikz, array, mathtools, bussproofs, stmaryrd, enumitem,
  stackengine, subcaption, cleveref}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\usepackage{xspace}
\usepackage[noend]{algpseudocode}

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

Primitive lenses include:
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get
  direction; insert it in the put direction.
  \item \kw{ins} $s$: insert the constant string $s$ in the get
  direction; delete it in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in
  both directions.
\end{itemize}
Such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, as well as lens composition.
Figure~\ref{fig:example-lens} illustrates the use of these
combinators to define a lens \cd{bib_to_end} that transforms data
in \bibtex{} to EndNote (and vice versa). 
\begin{figure}[t]
\begin{lstlisting}
let preamble : "@book{" $\Leftrightarrow$ "%0 Book\n%F " =
del "@book{" . ins "%0 Book\n%F "

let author_lens : (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author) =
del ",\nauthor = \""
. ins "\n%A "
. name
. (del " and "
. ins "\n"
. (ins "%A " . name . del " and " . ins "\n")*
. ins "%A "
. name
|| "")

let title_lens : ("\",\n" . bib_title . "},\n}") $\Leftrightarrow$ ("\n" . end_title) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{An example bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions, which are omitted for brevity.  
\ifdraft
\bcp{And a readability suggestion:
  define a named constant for a quote character.}
\fi}
\label{fig:example-lens}
\end{figure}

@@

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\begin{lstlisting}
let bib_to_end : bibtex $\Leftrightarrow$ endnote =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}, but first we discuss the \kw{normalize} combinator which
enjoys some special properties.

\subsection{The \kw{normalize} Combinator}
The \kw{normalize} combinator is special in two main ways, the first of
which has to do with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $Q$ is semantically
equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$; this is the
second property that distinguishes the {\em normalize} combinator from the other
QREs. The {\em normalize} combinator hence gives a sufficient condition for
potential QREs to be ``well-behaved'', particularly with respect to the
strategy for synthesizing quotient lenses which we shall introduce in
Section~\ref{synth}.

Having discussed the special properties of the \kw{normalize} combinator, we now
give the syntax and semantics of the other QRE combinators.
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &\id(R) \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function $\canonize(Q):\mathcal{L}(W(Q))
  \longrightarrow \mathcal{L}(K(Q))$ which given any $w \in \mathcal{L}(W(Q))$,
  computes $\canonize(Q)(w)$ as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \textit{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \textit{permutation} combinator allows for the string to match any
permutation the $Q_i$'s while also taking into account the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \textit{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

@@

The most complicated inference rule is the (Perm) rule for the
\textit{permutation} combinator. The second hypothesis for the
\textit{permutation} rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\textit{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$ which is
the kernel of the \textit{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

\begin{lstlisting}
let bib_to_end_q : bibtex $\Leftrightarrow$ endnote =  rquot (lquot(bibtex, bib_to_end), endnote)
\end{lstlisting}

@@

The Optician bijective lens synthesis algorithm works as follows. Given regular
expressions $R$ and $S$ and a set of example input--output pairs $\{(r_1,
s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to find a
bijective lens $\ell : R \Leftrightarrow S$ that agrees on the examples,
\textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and $s_i$ to
$r_i$ in the backward direction. The bijective lens synthesis algorithm is
guaranteed to succeed (eventually!) if such a bijective lens exists. There are
typically a large number of such lenses, so this algorithm chooses the one that
corresponds to a minimal \emph{alignment} of the regular expressions. Additional
details on this algorithm can be found in the original paper on synthesizing
bijective lenses~\cite{optician}.
